---
title: "Biostat 203B Homework 5"
subtitle: Due Mar 20 @ 11:59PM
author: "Julie Lee, 806409381"
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: false
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
---

Loading all necessary packages

```{r}
library(rsample)
library(tidymodels)
library(xgboost)
library(kernlab)
library(glmnet)  
library(xgboost)
library(doParallel)
library(pROC) 
library(corrr)
library(ggfocus)
library(caret)
library(e1071)
library(randomForest)
library(ranger)

```


## Predicting ICU duration

Using the ICU cohort `mimiciv_icu_cohort.rds` you built in Homework 4, develop at least three machine learning approaches (logistic regression with enet regularization, random forest, boosting, SVM, MLP, etc) plus a model stacking approach for predicting whether a patient's ICU stay will be longer than 2 days. You should use the `los_long` variable as the outcome. You algorithms can use patient demographic information (gender, age at ICU `intime`, marital status, race), ICU admission information (first care unit), the last lab measurements before the ICU stay, and first vital measurements during ICU stay as features. You are welcome to use any feature engineering techniques you think are appropriate; but make sure to not use features that are not available at an ICU stay's `intime`. For instance, `last_careunit` cannot be used in your algorithms. 

1. Data preprocessing and feature engineering.

```{r}
mimiciv_icu_cohort <- readRDS("../hw4/mimiciv_shiny/mimic_icu_cohort.rds")
```

**Data Preprocessing and Feature Engineering Explanation: ** 

We began our data preprocessing by creating a new binary variable, los_long, which is set to TRUE when a patient's length of stay (los) is greater than or equal to two days. To handle missing data, we converted categorical variables into factors and applied median imputation for missing numerical values and mode imputation for missing categorical values. For feature selection, we included patient demographic information (gender, age at ICU admission, marital status, race), ICU admission details (first care unit), the last recorded laboratory measurements prior to ICU admission, and the first recorded vital signs during the ICU stay—both of which were extracted from previous reports. To ensure consistency and improve model performance, we standardized all continuous variables. Additionally, categorical variables were converted into factors to facilitate proper encoding for machine learning models. This structured preprocessing ensures that the dataset is well-prepared for predictive modeling while avoiding data leakage by only incorporating information available at the time of ICU admission. Note that there are 18 total features used in the model to predict the whether a patient's lengt of stay will be greater than or equal to 2 days. Note that we will include subject_id, hadm_id, and stay_id for grading purposes only but will remove them as features for the model in future steps. 


```{r}
# Begin preprocessing by creating binary outcome variable 'los_long'
mimiciv_icu_cohort <- mimiciv_icu_cohort %>%
  filter(!is.na(los)) %>% 
  mutate(
    los_long = as.factor(as.integer(los >= 2))  # Convert directly to factor (1 = TRUE, 0 = FALSE)
  ) %>%
  select(subject_id, hadm_id, stay_id, los_long,
         gender, age_at_intime, marital_status, race, first_careunit,
         heart_rate, non_invasive_blood_pressure_systolic, non_invasive_blood_pressure_diastolic,
         respiratory_rate, temperature_fahrenheit, bicarbonate, glucose,
         potassium, sodium, chloride, creatinine, hematocrit, wbc) %>%
  
  # Convert categorical variables into factors and handle missing values using mode imputation
  mutate(across(c(gender, marital_status, race, first_careunit), 
                ~ replace_na(.x, names(sort(table(.x), decreasing = TRUE))[1])),
         across(c(gender, marital_status, race, first_careunit), factor)) %>%

  # Median imputation for numeric variables
  mutate(across(c(age_at_intime, heart_rate, non_invasive_blood_pressure_systolic,
                  non_invasive_blood_pressure_diastolic, respiratory_rate,
                  temperature_fahrenheit, bicarbonate, glucose, potassium,
                  sodium, chloride, creatinine, hematocrit, wbc),
                ~ if_else(is.na(.x), median(.x, na.rm = TRUE), .x))) %>%

  # Standardize continuous numerical variables
  mutate(across(c(age_at_intime, heart_rate, non_invasive_blood_pressure_systolic,
                  non_invasive_blood_pressure_diastolic, respiratory_rate,
                  temperature_fahrenheit, bicarbonate, glucose, potassium,
                  sodium, chloride, creatinine, hematocrit, wbc),
                ~ as.numeric(scale(.x))))


mimiciv_icu_cohort
```

2. Partition data into 50% training set and 50% test set. Stratify partitioning according to `los_long`. For grading purpose, sort the data by `subject_id`, `hadm_id`, and `stay_id` and use the seed `203` for the initial data split. Below is the sample code.

```{r}
set.seed(203)

mimiciv_icu_cohort <- mimiciv_icu_cohort %>%
  arrange(subject_id, hadm_id, stay_id)

# Perform stratified data splitting (50% train, 50% test)
data_split <- initial_split(mimiciv_icu_cohort, prop = 0.5, strata = los_long)

train_data <- training(data_split)
test_data <- testing(data_split)

# Remove subject_id, hadm_id, and stay_id from train and test sets
train_data <- train_data %>%
  select(-subject_id, -hadm_id, -stay_id)

test_data <- test_data %>%
  select(-subject_id, -hadm_id, -stay_id)

# Display the first few rows of train and test sets
head(train_data)
head(test_data)
```


3. Train and tune the models using the training set.

#Approach 1: Logistic Regression 

```{r}
set.seed(203)  # Ensure reproducibility

train_sampled <- train_data %>% sample_frac(0.20) 

# Convert categorical variables to dummy variables and remove the response variable
train_x <- model.matrix(los_long ~ . - 1, data = train_sampled)  # Remove intercept
train_y <- as.numeric(as.character(train_sampled$los_long))  # Convert factor to numeric (0,1)

test_x <- model.matrix(los_long ~ . - 1, data = test_data)
test_y <- as.numeric(as.character(test_data$los_long))

registerDoParallel(cores = parallel::detectCores() - 1)  
set.seed(123)

# Define grid of tuning parameters
alpha_values <- seq(0, 1, length.out = 25)  # Alpha (L1 ratio)
lambda_values <- 10^seq(-4, 1, length.out = 10)  # Lambda (regularization)

# Generate 250 (25 x 10) parameter combinations
tuning_grid <- expand.grid(alpha = alpha_values, lambda = lambda_values)

# Initialize storage for results
cv_results <- list()

# Iterate through each alpha-lambda combination
for (i in 1:nrow(tuning_grid)) {
  alpha_i <- tuning_grid$alpha[i]
  lambda_i <- tuning_grid$lambda[i]
  
  # Perform 5-fold cross-validation
  cv_fit <- cv.glmnet(x = train_x, 
                       y = train_y, 
                       family = "binomial",
                       alpha = alpha_i, 
                       lambda = lambda_values, 
                       nfolds = 5,
                       parallel = TRUE) 
  
  # Store results
  cv_results[[i]] <- list(alpha = alpha_i, lambda = cv_fit$lambda.min, error = min(cv_fit$cvm))
}

# Convert list to dataframe
cv_results_df <- do.call(rbind, lapply(cv_results, as.data.frame))

# Find the best combination (lowest cross-validation error)
best_params <- cv_results_df[which.min(cv_results_df$error), ]
best_alpha <- best_params$alpha
best_lambda <- best_params$lambda

print(best_params)
final_model <- glmnet(x = train_x, 
                      y = train_y, 
                      family = "binomial",
                      alpha = best_alpha, 
                      lambda = best_lambda)

# Predict on the test set
test_predictions <- predict(final_model, newx = test_x, type = "response")

# Convert probabilities to binary predictions
test_pred_labels <- ifelse(test_predictions > 0.5, 1, 0)

# Evaluate model performance
conf_matrix <- table(Predicted = test_pred_labels, Actual = test_y)
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

print(conf_matrix)
print(paste("Test Accuracy:", round(accuracy, 3)))

# Compute ROC Curve
roc_curve <- roc(test_y, test_predictions)  # test_predictions are probabilities

# Calculate AUC
auc_score <- auc(roc_curve)

# Print AUC Score
print(paste("AUC Score:", round(auc_score, 3)))

# Plot ROC Curve
plot(roc_curve, col = "blue", main = "ROC Curve for Logistic Regression with ENET")
abline(a = 0, b = 1, lty = 2, col = "red")  # Add diagonal line (random classifier)

```

#Approach 2: XG Boosting 
```{r}
set.seed(203)  # Ensure reproducibility

# Sample only 25% of the training data
train_sampled <- train_data %>% sample_frac(0.20)

# Convert categorical variables to dummy variables
train_x <- model.matrix(los_long ~ . - 1, data = train_sampled)  # Remove intercept
test_x <- model.matrix(los_long ~ . - 1, data = test_data)

# Convert target variable to numeric format (XGBoost requires numeric encoding)
train_y <- as.numeric(as.character(train_sampled$los_long))  # 0 or 1
test_y <- as.numeric(as.character(test_data$los_long))    # 0 or 1


# Convert to XGBoost DMatrix format
dtrain <- xgb.DMatrix(data = train_x, label = train_y)
dtest <- xgb.DMatrix(data = test_x, label = test_y)


registerDoParallel(cores = parallel::detectCores() - 1)  # Use all but one core

set.seed(123)

# Define tuning grid (21 combinations)
tuning_grid <- expand.grid(
  eta = seq(0.01, 0.3, length.out = 3),  # Learning rate
  max_depth = c(3, 5, 7),  # Tree depth
  min_child_weight = c(1, 3, 5),  # Minimum sum of instance weight (hessian)
  subsample = c(0.6, 0.8, 1),  # Subsample ratio
  colsample_bytree = c(0.6, 0.8, 1)  # Column sample by tree
)

# Randomly sample 21 parameter combinations
tuning_grid <- tuning_grid[sample(1:nrow(tuning_grid), 21), ]


cv_results <- list()

for (i in 1:nrow(tuning_grid)) {
  params <- list(
    booster = "gbtree",
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = tuning_grid$eta[i],
    max_depth = tuning_grid$max_depth[i],
    min_child_weight = tuning_grid$min_child_weight[i],
    subsample = tuning_grid$subsample[i],
    colsample_bytree = tuning_grid$colsample_bytree[i]
  )

  # 5-fold cross-validation
  cv_model <- xgb.cv(params = params, data = dtrain, nrounds = 100, 
                      nfold = 5, early_stopping_rounds = 10, maximize = TRUE, 
                      verbose = FALSE, parallel = TRUE)

  # Store results
  best_iteration <- cv_model$best_iteration
  best_auc <- max(cv_model$evaluation_log$test_auc_mean)
  
  cv_results[[i]] <- c(tuning_grid[i, ], best_iteration = best_iteration, AUC = best_auc)
}

# Convert results list to dataframe
cv_results_df <- do.call(rbind, lapply(cv_results, as.data.frame))

# Find the best combination based on AUC
best_params <- cv_results_df[which.max(cv_results_df$AUC), ]
print(best_params)

final_params <- list(
  booster = "gbtree",
  objective = "binary:logistic",
  eval_metric = "auc",
  eta = best_params$eta,
  max_depth = best_params$max_depth,
  min_child_weight = best_params$min_child_weight,
  subsample = best_params$subsample,
  colsample_bytree = best_params$colsample_bytree
)

final_model <- xgb.train(params = final_params, data = dtrain, 
                         nrounds = best_params$best_iteration, 
                         watchlist = list(train = dtrain, test = dtest), 
                         early_stopping_rounds = 10, verbose = FALSE)

# Predict probabilities
test_predictions <- predict(final_model, newdata = dtest)

# Optimize threshold using Youden’s Index
roc_curve <- roc(test_y, test_predictions)
best_threshold <- coords(roc_curve, "best", ret = "threshold")

# Convert probabilities to binary predictions
test_pred_labels <- ifelse(test_predictions > best_threshold, 1, 0)

print(conf_matrix)
print(paste("Test Accuracy:", round(accuracy, 3)))

# Compute AUC
auc_score <- auc(roc_curve)
print(paste("AUC Score:", round(auc_score, 3)))

# Plot ROC Curve
plot(roc_curve, col = "blue", main = "ROC Curve for XGBoost Model")
abline(a = 0, b = 1, lty = 2, col = "red")  # Add diagonal reference line
```

#Approach 3: 

```{r}
set.seed(203)

# Reduce training size to 15% for faster computation
train_sampled <- train_data %>% sample_frac(0.15)

# Convert categorical variables to factors
train_sampled <- train_sampled %>%
  mutate(across(where(is.character), as.factor))

test_data <- test_data %>%
  mutate(across(where(is.character), as.factor))

# Ensure target variable is a factor
train_sampled$los_long <- factor(train_sampled$los_long, levels = c(0, 1), labels = c("Class0", "Class1"))
test_data$los_long <- factor(test_data$los_long, levels = c(0, 1), labels = c("Class0", "Class1"))

# Parallel Processing
registerDoParallel(cores = parallel::detectCores() - 1)

set.seed(123)

# Define tuning grid with `mtry`, `splitrule`, and `min.node.size`
tuning_grid <- expand.grid(
  mtry = seq(2, floor(log2(ncol(train_sampled) - 1)), by = 2),  # Selects optimal features per split
  splitrule = c("gini", "extratrees"),  # Different splitting criteria
  min.node.size = c(1, 5, 10)  # Minimum observations per terminal node
)

# Use 3-Fold Cross-Validation for Faster Training
cv_control <- trainControl(method = "cv", number = 3,  
                           summaryFunction = twoClassSummary, 
                           classProbs = TRUE, 
                           allowParallel = TRUE)

# Train Random Forest with `ranger` for speed
rf_model <- train(los_long ~ ., data = train_sampled,
                  method = "ranger",  # Use ranger instead of randomForest
                  trControl = cv_control,
                  tuneGrid = tuning_grid,
                  metric = "ROC")

# Print best hyperparameters
print(rf_model$bestTune)

# Train Final Model with Optimized Parameters
final_rf <- ranger(los_long ~ ., data = train_sampled, 
                   mtry = rf_model$bestTune$mtry, 
                   splitrule = rf_model$bestTune$splitrule,
                   min.node.size = rf_model$bestTune$min.node.size,
                   num.trees = 200,  # Reduce trees for speed
                   importance = "impurity", 
                   probability = TRUE, 
                   num.threads = parallel::detectCores() - 1)  # Use all CPU cores

# Predict probabilities
test_predictions <- predict(final_rf, test_data, type = "response")$predictions[,2]

# Optimize threshold using Youden’s Index
roc_curve <- roc(test_data$los_long, test_predictions)
best_threshold <- coords(roc_curve, "best", ret = "threshold")

# Convert probabilities to binary predictions
test_pred_labels <- ifelse(test_predictions > best_threshold, "Class1", "Class0")

# Confusion Matrix & Performance Metrics
conf_matrix <- table(Predicted = test_pred_labels, Actual = test_data$los_long)
accuracy <- sum(diag(conf_matrix)) / sum(conf_matrix)

print(conf_matrix)
print(paste("Test Accuracy:", round(accuracy, 3)))

# Compute AUC
auc_score <- auc(roc_curve)
print(paste("AUC Score:", round(auc_score, 3)))

# Plot ROC Curve
plot(roc_curve, col = "blue", main = "ROC Curve for Optimized Random Forest")
abline(a = 0, b = 1, lty = 2, col = "red")  # Add diagonal reference line

# Feature Importance
importance(final_rf)


```





4. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each machine learning algorithm and the model stacking. Interpret the results. What are the most important features in predicting long ICU stays? How do the models compare in terms of performance and interpretability?

Model #1: 

Model #2: XG Boosting 

Model #3: 


1. 




